---
title: "individual1.Rmd"
author: "Taylor Turner"
date: "9/23/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, inlcude=FALSE}
library(dplyr)
library(readxl)
library('ggplot2')
dat <- readxl::read_xls('./GSS.xls')
dat <- dat%>% 
  rename(year = 'Gss year for this respondent', happiness = 'General happiness', partyid = 'Political party affiliation', income = 'Total family income', sex = 'Respondents sex', educ = 'Highest year of school completed', age = 'Age of respondent', marital = 'Marital status', id = 'Respondent id number', ballot = 'Ballot used for interview')
dat$educ <- suppressWarnings(as.numeric(dat$educ))
dat$age <- suppressWarnings(as.numeric(dat$age))
```

1. We can see that for all party identifications, the average years of school is from 12 to 14 years of schooling. The political party identification with the greatest average years of school is "strong republican." The political party identification with the smallest average years of school is "don't know".
```{r}
dat%>%
  select(partyid, educ)%>%
  filter(educ<21)%>%
  group_by(partyid)%>%
  summarise(educ_mean = mean(educ), .groups='drop')
```

2.
```{r}
df <- dat%>%
  group_by(partyid)%>%
  filter(educ<21)%>%
  mutate(partyMedian = median(educ))%>%
  filter(educ > partyMedian)%>%
  select(-partyMedian)%>%
  ungroup()

str(df)
```

3. From the plot, we can see that "married" has the most happy respondents. The group with the least amount of happy people are those who did not answer. However, if we disregard the "no answer" category, the group with the least amount of happy people are those who are separated.
For this I used the variables happiness and marital.
```{r}
dat%>% 
  select(happiness, marital)%>% 
  ggplot(aes(x=marital, fill=happiness)) + geom_bar(position='fill') + ggtitle("Happiness and Marital Status")

dat%>% 
  select(happiness, marital)%>% 
  group_by(marital)%>%
  count(happiness)
```

4. It doesn't seem like political party has much of an effect on happiness and marital status. For all political party identifications, married has the most happy respondents than the rest. Separated still had the least amount of happy respondents. For this I used the variables happiness, partyid, and marital.
```{r}
ggplot(dat, aes(x=marital, fill=happiness)) + geom_bar() + facet_wrap(~partyid, scales="free_x") + coord_flip() + ggtitle("Happiness, Marital Status, and Political Party") + theme(axis.text.x = element_text(angle = 90))

dat%>%
  select(happiness, partyid, marital)%>%
  group_by(partyid, marital)%>%
  count(happiness)
```

5. There doesn't seem to be clear patterns in the correlation between years of education and marital status. For this I used the variables educ and marital. There looks to be less widowed respondents with more years of education. There also is less divorces with respondents that have less years of education. There are more respondents that were never married at the years of education goes up. This could be due to the respondent focusing more of their time on school than finding a parter.
```{r}
dat%>% 
  select(educ, marital)%>% 
  filter(educ < 21)%>%
  ggplot(aes(x=educ, fill=marital)) + geom_bar(position='fill') + ggtitle("Education and Marital Status") + scale_x_continuous(breaks = 0:20, "Years of Education")

dat%>% 
  select(educ, marital)%>% 
  filter(educ < 21)%>%
  group_by(marital)%>%
  count(educ)
```

6. My first question is "how does the respondent's sex affect the happiness and marital status?" This question is similar to question 4, but replacing political party with gender. We can see that respondents that are married are still the happiest, regardless of gender. Married women have the most happy repsonses. For this I used the variables happiness, sex, and marital.
```{r}
ggplot(dat, aes(x=marital, fill=happiness)) + geom_bar() + facet_wrap(~sex, scales="free_x") + ggtitle("Happiness, Marital Status, and Sex") + theme(axis.text.x = element_text(angle = 90))

dat%>%
  select(happiness, sex, marital)%>%
  group_by(sex, marital)%>%
  count(happiness)
```

My second question is "how does the respondent's age affect their happiness?" We can see that happiness levels are generally pretty even across all age groups and there don't seem to be any obvious patterns. Also, the majority of respondents in all age groups are happy. For this I used the variables happiness and age. 
```{r}
dat%>% 
  select(happiness, age)%>% 
  filter(age > 17)%>% 
  ggplot(aes(x=age, fill=happiness)) + geom_bar(position='fill') + ggtitle("Happiness and Age")

dat%>%
  select(happiness, age)%>%
  filter(age < 90)%>%
  count(age, happiness)
```